\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2009-08-13 papers/jss/article.tex
- more changes to jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@309 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and e.g.\ \citet{Schmittmann2006} for a recent
application.  In economics, latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and
\citealp{Ghysels1994}).  Further applications include speech
recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
genetics \citep{Krogh1998}.  In those latter areas of application,
latent Markov models are usually referred to as hidden Markov models.
For more examples of applications, see e.g.\
\citet[][chapter~1]{Cappe2005}.

The \pkg{depmixS4} package was motivated by the fact that while Markov
models are used commonly in the social sciences, no comprehensive
package was available for fitting such models.  Common software for
estimating Markovian models include Panmark \citep{Pol1996}, and for
latent class models Latent Gold \citep{Vermunt2003}.  Those programs
are lacking a number of important features, besides not being freely
available.  There are currently some packages in \proglang{R} that
handle hidden Markov models but they lack a number of features that we
needed in our research.  In particular, \pkg{depmixS4} was designed to
meet the following goals:

@@

	\item to be able to include covariates in the prior or initial
	state probabilities of models

	\item to be easily extensible, in particular, to allow users to
	easily add new uni- or multivariate response distributions, and
	similarly for the addition of other transition models, e.g.,
	continuous time observation models

@@

Although \pkg{depmixS4} was designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case when
$T=1$.  In this case, there are no time dependencies between
observed data and the model reduces to a finite mixture or
latent class model. Although there are other specialized packages to
deal with mixture data, as far as we know these 
don't allow the inclusion of covariates on the prior probabilities of
class membership. The possibility to estimating the effects of covariates on 
prior and transition probabilities is a distinguishing feature of 
\pkg{depmixS4}. In the next section, an outline is provided of the model and 
the likelihood equations.

@@

The data considered here have the general form $\vc{O}_{1:T}=O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
participant in a response time experiment. The data consists of three
variables, response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accuracy responding. 
These variables are measured on 168, 134 and 137 occasions respectively (in
Figure~\ref{fig:speed} the first part of this series is plotted). These data
are more fully described in \citet{Dutilh2009}.

@@

The latent Markov model is commonly associated with data of this type,
although usually only multinomial response variables are considered.
However, common estimation procedures, such as those implemented in
\citet{Pol1996}, are not suitable for long time series due to
underflow problems.  In contrast, the hidden Markov model is typically
only used for `long' univariate time series \citep[][,
chapter~1]{Cappe2005}.  In the next paragraphs, the likelihood
equation and estimation procedures for the dependent mixture model are
described for data of the above form.  We use the term ``dependent
mixture model'' because one of the authors (Ingmar Visser) thought it
was time for a new name for these models\footnote{Only later did I
find out that \citet{Leroux1992} already coined the term dependent
mixture models in an application with hidden Markov mixtures of
Poisson count data.}.

The likelihood of the dependent mixture model conditional on the
unknown (or hidden) state sequence $S_{1:T}$ and the model is given
by:
\begin{equation}
	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
\end{equation}
where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
at the likelihood of the data given the parameter vector
$\greekv{\theta}$, i.e.\ without the state sequence, we need to sum
above likelihood over all possible state sequences. This likelihood 
is written as: 
\begin{equation}
	\Prob(\vc{O}_{1:T}|\vc{S}_{1:T},\greekv{\theta}) 
	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
	\prod_{t=1}^{T-1} a_{ij}(z_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
\end{equation}
where we have the following elements:
\begin{enumerate}

	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
	of $n$ latent classes or states; we write for short $S_{i}$ to
	denote $S_{t}=i$.

@@

	\item $\vc{b}_{j}$ is a vector of observation densities
	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
	conditional densities of observations $O_{t}^k$ associated with
	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
	$k=1, \ldots, m$. 
\end{enumerate}

@@

In the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli for the
accuracy data.  In the models we are considering here, both the
transition probability functions $a_{ij}$and the initial state
probability functions $\greekv{\pi}$ may depend on covariates as well
as the response distributions $\vc{b_{j}^{k}}$.  See for example
\citet{Fruhwirth2006} for an overview of hidden Markov models with
extensions.

@@

The log-likelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
	Pr(\vc{O}_{t}|\vc{O}_{1:t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= \Prob(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
parameters are estimated by iteratively maximising the 
expected joint likelihood of the parameters given the observations and states. 
Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
be the general parameter vector consisting of three subvectors with parameters 
for the prior model, transition model, and response model respectively. The 
joint log likelihood can be written as
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $O_{1:T}$. The expected 
log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\ln \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
right hand side of (\ref{eq:Q}) consists of three separate parts, we can 
maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and 
$\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and 
$\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in the 
\pkg{nnet} package \citep{Venables2002}, and maximisation for 
$\greekv{\theta}_3$ by the standard \code{glm} routine. Note that for the latter 
maximisation, the expected values $\gamma_t(j)$ are used as prior weights of the 
observations $O^k_t$.

@@

Throughout this article a data set called \code{speed} is used.  It
consists of three time series with three variables: response time,
accuracy, and a covariate Pacc which defines the relative pay-off for
speeded and accurate responding.  The participant in this experiment
switches between fast responding at chance level and relatively slower
responding at a high level of accuracy.  Interesting hypotheses to
test are: is the switching regime symmetric?  Is there evidence for
two states or does one state suffice?  Is the guessing state actually
a guessing state, i.e., is the probability of a correct response at
chance level (0.5)?

@@

A dependent mixture model is defined by the number of states, and by
the response distribution functions, and can be created with the
\code{depmix}-function as follows:

@@

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log likelihood and the AIC and BIC
values:
\begin{CodeChunk}
\begin{CodeInput}
> fm
\end{CodeInput}
\begin{CodeOutput}
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34175 (df=7)
AIC:  182.6835 
BIC:  211.275 
\end{CodeOutput}
\end{CodeChunk}
These statistics can also be extracted using \code{logLik},
\code{AIC} and \code{BIC}, respectively.

@@

@@

The transition probabilities and the initial state probabilities
are parameterized using the multinomial
logistic model.   In particular, each row of the transition matrix is
parameterized by a baseline category logistic multinomial, 
meaning that the parameter for the
base category is fixed at zero \citep[see][p.\ 267
ff., for multinomial logistic models and various
parameterizations]{Agresti2002}. See also \citet{Chung2007} for similar models, 
latent transition models using logistic regression on the transition parameters. 
They fit such models on repeated measurement
data ($T=2$) using Bayesian methods.  The default baseline category is the 
first state.
Hence, for example, for a 3-state model, the initial state probability
model would have three parameters of which the first is fixed at zero
and the other two are freely estimated.

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
  transition=~scale(Pacc), instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

@@

To see the ordering of parameters use the following:
\begin{CodeChunk}
\begin{CodeInput}
> mod <- setpars(mod, value=1:npar(mod))
> mod
\end{CodeInput}
\end{CodeChunk}

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
\begin{CodeChunk}
\begin{CodeInput}
> mod <- setpars(mod, getpars(mod,which="fixed"))
> mod
\end{CodeInput}
\end{CodeChunk}

When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
\begin{CodeChunk}
\begin{CodeInput}
> trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
> mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	  nstates=2,family=list(gaussian(),multinomial()),
	  trstart=trst,inst=c(.999,0.001))
> fm1 <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
After this, we use the fitted values from this model to constrain the 
regression coefficients on the transition matrix (parameters numbers~6 and~10):
\begin{CodeChunk}
\begin{CodeInput}
# start with fixed and free parameters
> conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
> conpat[6] <- conpat[10] <- 2
> fm2 <- fit(fm,equal=conpat)
\end{CodeInput}
\end{CodeChunk}
Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. The function
\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
associated $p$-value with appropriate degrees of freedom for testing the
tenability of constraints \citep{Dannemann2007}. Note that these arguments 
provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

Only the constructor function and the fit function are provided here
and the complete code is provided in the helpfile for \code{makeDepmix}. 
The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} packages \cite{Stasinopoulos2009a,Stasinopoulos2009b} for 
computing the \code{dens}ity and for \code{fit}ting the parameters.

>>>>

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes (see \citealp{Wickens1982},
for an overview, and e.g. \citealp{Schmittmann2006}, for a recent
application).  In economics, latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and
\citealp{Ghysels1994}).  Further applications include speech
recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
genetics \citep{Krogh1998}.  In these latter areas of application,
latent Markov models are usually referred to as hidden Markov models. 
See for example \citet{Fruhwirth2006} for an overview of hidden Markov models 
with extensions. Further examples of applications can be found in e.g.\
\citet[][chapter~1]{Cappe2005}.

The \pkg{depmixS4} package was motivated by the fact that while Markov
models are used commonly in the social sciences, no comprehensive
package was available for fitting such models.  Existing software for
estimating Markovian models include Panmark \citep{Pol1996}, and for
latent class models Latent Gold \citep{Vermunt2003}. These programs
lack a number of important features, besides not being freely
available. There are currently some packages in \proglang{R} that
handle hidden Markov models but they lack a number of features that we
needed in our research. In particular, \pkg{depmixS4} was designed to
meet the following goals:

@@

	\item to be able to include covariates in the prior or initial
	state probabilities

	\item to be easily extensible, in particular, to allow users to
	easily add new uni- or multivariate response distributions and
	new transition models, e.g., continuous time observation models

@@

Although \pkg{depmixS4} was designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case when
$T=1$.  In this case, there are no time dependencies between
observed data and the model reduces to a finite mixture or
latent class model. While there are specialized packages to
deal with mixture data, as far as we know these 
don't allow the inclusion of covariates on the prior probabilities of
class membership. The possibility to estimate the effects of covariates on 
prior and transition probabilities is a distinguishing feature of 
\pkg{depmixS4}. In the next section, we provide an outline of the model and 
likelihood equations.

@@

The data considered here have the general form $\vc{O}_{1:T}= (O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m})$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
participant in a psychological response time experiment. The data consists of three
variables, response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accurate responding. 
These variables are measured on 168, 134 and 137 occasions respectively (the 
first part of this series is plotted in
Figure~\ref{fig:speed}). These data
are more fully described in \citet{Dutilh2009}.

@@

The latent Markov model is usually associated with data of this type,
in particular for multinomially distributed responses. % response variables are considered.
However, commonly employed estimation procedures
\citep[e.g.,][]{Pol1996}, are not suitable for long time series due to
underflow problems. In contrast, the hidden Markov model is typically
only used for `long' univariate time series \citep[][chapter~1]{Cappe2005}. 
We use the term ``dependent mixture model'' because one of the authors 
(Ingmar Visser) thought it
was time for a new name to relate these models\footnote{Only later did I
find out that \citet{Leroux1992} already coined the term dependent
mixture models in an application with hidden Markov mixtures of
Poisson count data.}.

The fundamental assumption of a dependent mixture model is that at any time 
point, the observations are distributed as a mixture with $n$ components 
(or states), and that time-dependencies between the observations are due to 
time-dependencies between the mixture components (i.e., transition probabilities
between the components). These latter dependencies are assumed to
follow a first-order Markov process. In the models we are considering here, the 
mixture distributions, the initial mixture probabilities and transition 
probabilities can all dependent on covariates $\vc{z}_t$.

%transition probability functions $a_{ij}$ and the initial state
%probability functions $\greekv{\pi}$ may depend on covariates as well
%as the response distributions $\vc{b_{j}^{k}}$.  See for example
%\citet{Fruhwirth2006} for an overview of hidden Markov models with
%extensions.

In a dependent mixture model, the joint likelihood of observations $\vc{O}_{1:T}$ and 
latent states $\vc{S}_{1:T} = (S_1,\ldots,S_T)$, given model parameters 
$\greekvec{\theta}$ and covariates $\vc{z}_{1:T} = (\vc{z}_1,\ldots,\vc{z}_T)$, 
can be written as
\begin{equation}
	\Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\greekv{\theta},\vc{z}_{1:T}) =  
	\pi_{i}(\vc{z}_1) \vc{b}_{S_t}(\vc{O}_1|\vc{z}_{1})
	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{S_t}(\vc{O}_{t+1}|\vc{z}_{t+1}),
\end{equation}
where we have the following elements:
\begin{enumerate}

	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
	of $n$ latent classes or states. %; we write for short $S_{i}$ to
	%denote $S_{t}=i$.

@@

	\item $\vc{b}_{S_t}$ is a vector of observation densities
	$b_{j}^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
	conditional densities of observations $O_{t}^k$ associated with
	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
	$k=1, \ldots, m$. 
\end{enumerate}

%In the next paragraphs, the likelihood
%equation and estimation procedures for the dependent mixture model are
%described for data of the above form.

%The likelihood of the dependent mixture model conditional on the
%unknown (or hidden) state sequence $\vc{S}_{1:T} = (S_1,\ldots,S_T)$ 
%and the model is given
%by:
%\begin{equation}
%	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
%	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
%\end{equation}
%where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
%at the marginal likelihood of the data for parameter vector
%$\greekv{\theta}$, i.e.\ unconditional on the state sequence, we need to sum
%above likelihood over all possible state sequences. This likelihood 
%is written as: 
%\begin{equation}
%	\Prob(\vc{O}_{1:T}|\greekv{\theta}) =  
%	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
%	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
%\end{equation}
%where we have the following elements:
%\begin{enumerate}
%	
%	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
%	of $n$ latent classes or states; we write for short $S_{i}$ to
%	denote $S_{t}=i$.
%	
%	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
%	giving the probability of class/state $i$ at time $t=1$ with
%	covariate $\vc{z}_1$.
%	
%	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
%	provides the probability of a transition from state $i$ to state
%	$j$ with covariate $\vc{z}_t$,
%	
%	\item $\vc{b}_{j}$ is a vector of observation densities
%	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
%	conditional densities of observations $O_{t}^k$ associated with
%	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
%	$k=1, \ldots, m$. 
%\end{enumerate}

@@

For the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli distribution for the
accuracy variable.  In the models we are considering here, both the
transition probability functions $a_{ij}$ and the initial state
probability functions $\greekv{\pi}$ may depend on covariates as well
as the response distributions $b_{j}^{k}$.  See for example
\citet{Fruhwirth2006} for an overview of hidden Markov models with
extensions.

@@

To obtain maximum likelihood estimates of the model parameters, we need the
marginal likelihood of the observations. For hidden Markov models, this 
marginal (log-)likelihood is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm. \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time. They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters and covariates is dropped here):
\begin{equation}
	L_{T} = \Prob(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
	\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}), 
	\label{condLike}
\end{equation}
where $\Prob(\vc{O}_{1}|\vc{O}_{0}):=\Prob(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$\Prob(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=\Prob(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= \Prob(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
%\begin{split}
	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1:(t-1)}) %\\
	= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
%\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
parameters are estimated by iteratively maximising the 
expected joint likelihood of the parameters given the observations and states. 
Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
be the general parameter vector consisting of three subvectors with parameters 
for the prior model, transition model, and response model respectively. The 
joint log-likelihood can be written as
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $O_{1:T}$. The expected 
log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\ln \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
forward-backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
right hand side of (\ref{eq:Q}) consists of three separate parts, we can 
maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and 
$\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and 
$\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in the 
\pkg{nnet} package \citep{Venables2002}, and maximisation for 
$\greekv{\theta}_3$ by the standard \code{glm} routine. Note that for the latter 
maximisation, the expected values $\gamma_t(j)$ are used as prior weights of the 
observations $O^k_t$.

@@

Throughout this article a data set called \code{speed} is used.  As already 
indicated in the Introduction, it
consists of three time series with three variables: response time,
accuracy, and a covariate Pacc which defines the relative pay-off for
speeded and accurate responding.  The participant in this experiment
switches between fast responding at chance level and relatively slower
responding at a high level of accuracy.  Interesting hypotheses to
test are: is the switching regime symmetric?  Is there evidence for
two states or does one state suffice?  Is the guessing state actually
a guessing state, i.e., is the probability of a correct response at
chance level (0.5)?

@@

A dependent mixture model is defined by the number of states and the initial 
state, state transition, and response distribution functions. A default 
dependent mixture model can be created with the \code{depmix}-function as 
follows:

@@

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log-likelihood and the AIC and BIC
values:
\begin{CodeChunk}
\begin{CodeInput}
> fm
\end{CodeInput}
\begin{CodeOutput}
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34175 (df=7)
AIC:  182.6835 
BIC:  211.275 
\end{CodeOutput}
\end{CodeChunk}
These statistics can also be extracted using \code{logLik},
\code{AIC} and \code{BIC}, respectively.

@@

Note that, since no further arguments were specified, the initial state, 
state transition and response distributions were set to their defaults 
(multinomial distributions for the first two, and Gaussian distributions for the
response distributions).

@@

By default, the transition probabilities and the initial state probabilities
are parameterized using the multinomial
logistic model. More precisely, each row of the transition matrix is
parameterized by a baseline category logistic multinomial, 
meaning that the parameter for the
base category is fixed at zero \citep[see][p.\ 267
ff., for multinomial logistic models and various
parameterizations]{Agresti2002}. The default baseline category is the 
first state. Hence, for example, for a 3-state model, the initial state probability
model would have three parameters of which the first is fixed at zero
and the other two are freely estimated.

The multinomial logistic model allows us to include covariates 
on the initial state and transition probabilities. \citet{Chung2007} discuss a 
related latent transition model for repeated measurement
data ($T=2$) using logistic regression on the transition 
parameters; they rely on Bayesian methods of estimation. 
Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
    transition=~scale(Pacc), instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

@@

To see the ordering of parameters use the following:
\begin{CodeChunk}
\begin{CodeInput}
> setpars(mod, value=1:npar(mod))
\end{CodeInput}
\end{CodeChunk}

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
\begin{CodeChunk}
\begin{CodeInput}
> setpars(mod, getpars(mod,which="fixed"))
\end{CodeInput}
\end{CodeChunk}

When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
\begin{CodeChunk}
\begin{CodeInput}
> trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
> mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	  nstates=2,family=list(gaussian(),multinomial()),
	  trstart=trst,inst=c(.999,0.001))
> fm1 <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
After this, we use the fitted values from this model to constrain the 
regression coefficients on the transition matrix (parameters numbers~6 and~10):
\begin{CodeChunk}
\begin{CodeInput}
# start with fixed and free parameters
> conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
> conpat[6] <- conpat[10] <- 2
> fm2 <- fit(fm,equal=conpat)
\end{CodeInput}
\end{CodeChunk}
Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. The function
\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
associated $p$-value with appropriate degrees of freedom for testing the
tenability of constraints \citep{Dannemann2007}. Note that these arguments 
provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a Gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

Only the constructor and the fit method are provided here; the complete code 
can be found in the helpfile for \code{makeDepmix}. 
The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} 
packages \citep{Stasinopoulos2009a,Stasinopoulos2009b} for 
computing the \code{dens}ity and for \code{fit}ting the parameters.

The constructor method is defined as: 
\begin{CodeChunk}
\begin{CodeInput}
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
  standardGeneric("exgaus"))
